Problem 1

```{r}
#Importing the dataset
car<-data.frame(mtcars)
```

```{r}
#Check Structure of dataset
str(car)
```
```{r}
library(caret)
set.seed(200)
partition <- createDataPartition(car$am,times=1,p=0.8,list = F)
train <- car[partition,]
test <- car[-partition,]
```
```{r}
#fiitting a linear model
model <- lm(mpg~.,data=train)

#MSE on test set
mean((predict(model,test)-test$mpg)^2)
summary(model)
coef(model)
```
Only Attribute "wt" is relevant


Ridge Regression
```{r}
# Loading the library
library(glmnet)
```

```{r}
# Getting the independent variable
x <- model.matrix(mpg~.,train)[,-1]


# Getting the dependent variable
y <- train$mpg
```


Cross Validation using GLMNET
```{r}
# Setting the range of lambda values
lambda_seq <- 10^seq(5,-5,by = -.1)

# Using cross validation glmnet
ridge_cv <- cv.glmnet(x, y, alpha = 0,lambda = lambda_seq)
plot(ridge_cv)

```


```{r}
#Best lambda value
best_lambda <- ridge_cv$lambda.min
best_lambda

```

```{r}
# Building the Ridge Regression Model using GLMNET
fit <- glmnet(x, y, alpha = 0, lambda  = best_lambda)
```


```{r}
summary(fit)

```


```{r}
coef(ridge_cv,s="lambda.min")
```


```{r}
# Test Dataset
x1 =  model.matrix(mpg~.,test)[,-1]
model_predict <- predict(fit,s =,newx = x1, type = "response")

#MSE on test data
mean((model_predict-test$mpg)^2)
```
We can see that MSE on test data will decreases from 10.71 to 1.18 by
performing Ridge Regression.
As we can see after Ridge Regression the coefficients have  shrunk and are more close to zero but none of them are perfect zero.
Hence Ridge Regression has performed shrinkage.






Problem 2

```{r}
library(ggplot2)
library(lattice)
library(caret)
#Importing the dataset
data <- data.frame(swiss)

```


```{r}

#80-20 split using createDataPartition
set.seed(150)
partition <- createDataPartition(data$Fertility,p=0.8,list = F)
train <- data[partition,]
test <- data[-partition,]

```


```{r}
#fitting a linear fit
model <- lm(Fertility~.,train)

summary(model)

```
Agriculture, Examination, Catholic and Infant Mortality are relevant feature with coefficients as -0.17497, -0.05176,
0.11713, 1.03247




```{r}
#calculating test mse
mean((test$Fertility-predict(model,test))^2)
```

Lasso Regression
```{r}
# Loaging the library
library(Matrix)
library(foreach)
library(glmnet)
# Getting the independent variable
x <- model.matrix(Fertility~.,train)[,-1]

# Getting the dependent variable
y <- train$Fertility
```

Cross Validation Lasso GLMNET
```{r}
# Setting the range of lambda values
lambda_seq <- 10^seq(5,-5,by = -.1)

# Using cross validation glmnet
lasso_cv <- cv.glmnet(x, y, alpha = 1,lambda = lambda_seq)
plot(lasso_cv)

```


```{r}
#Best lambda value
best_lambda <- lasso_cv$lambda.min
best_lambda

```

```{r}
# Using glmnet function to build the ridge regression model
fit <- glmnet(x, y, alpha = 1, lambda  = best_lambda)

# Checking the model
summary(fit)

```


```{r}
#for testdata
x2 =  model.matrix(Fertility~.,test)[,-1]
model_predict <- predict(fit,s =,newx = x2, type = "response")
```


```{r}
#MSE on test data
mean((model_predict-test$Fertility)^2)
```


```{r}
#coefficients
coef(model)
coef(lasso_cv)
```
Compared to Linear fit Lasso Regularization has shrinked the coefficients and two of them are shrinked to
zero.



Problem 3
```{r}
concrete <- read.csv("D:\\Temp\\Concrete_Data.csv")
summary(concrete)
```

Changing the Column names
Taking Columns C1-C6
```{r}
colnames(concrete) = c("cem", "bfs", "fa", "water", "sp", "cagg", "fagg", "age", "ccs")
keeps = c("cem", "bfs", "fa", "water", "sp", "cagg", "ccs")
concrete = concrete[keeps]
summary(concrete)
```

```{r}
library(corrplot)
corrplot(cor(concrete), method = "number")
```

```{r}

library(mgcv)
model1 <- gam(ccs ~ cem + bfs + fa + water + sp + cagg , data=concrete)
summary(model1)

```
It appears we have statistical effects for CEM, BFS, but not for CAGG and the adjusted R-squared suggests a notable amount of the variance.

Using Smoothing Function
```{r}
model2 <- gam(ccs ~ s(cem) + s(bfs) + s(fa) + s(water) + s(sp) + s(cagg) , data=concrete)
summary(model2)
```

We can also note that this model accounts for much of the variance in CCS , with an adjusted R-squared of .531 . In short, it looks like the CEM is associated with CCS.



```{r}
model1.sse <- sum(fitted(model1)-concrete$ccs)^2
model1.ssr <- sum(fitted(model1) -mean(concrete$ccs))^2
model1.sst = model1.sse + model1.ssr

rsqr_main=1-(model1.sse/model1.sst)
print(rsqr_main)
```
```{r}
model2.sse <- sum(fitted(model2)-concrete$ccs)^2
model2.ssr <- sum(fitted(model2) -mean(concrete$ccs))^2
model2.sst = model2.sse + model2.ssr

rsqr_sm=1-(model2.sse/model2.sst)
print(rsqr_sm)
```


Comparison of Model
```{r}
anova(model1, model2, test="Chisq")

```

We couldn’t have assumed as such already, but now we have additional statistical evidence to suggest that incorporating nonlinear relationships of the covariates improves the model.


Visualizing with Visreg Library
```{r}
library(visreg)
visreg(model1,'cem')
visreg(model2,'cem') 
```
The result is a plot of how the expected value of the CCS changes as a function of x (CEM), with all other variables in the model held fixed.
It includes 
(1) the expected value (blue line) 
(2) a confidence interval for the expected value (gray band)
(3) partial residuals (dark gray dots).



```{r}
visreg(model1,'bfs') 
visreg(model2,'bfs') 
```

```{r}
visreg(model1,'fa') 
visreg(model2,'fa')
```

```{r}
visreg(model1,'water') 
visreg(model2,'water')
```

```{r}
visreg(model1,'sp') 
visreg(model2,'sp')
```


```{r}
visreg(model1,'cagg') 
visreg(model2,'cagg')
```



From CEM  graph we can see that, the confidence interval after applying smoothing function has greater value as compared to the model before smoothing function.
After applying the smoothing function, the confidence interval gets better.

















